\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-09-04 writing/manuscript.Rmd
started editing the discussion

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles. Consequently, it is still unclear how mammals will adapt their movement behaviour in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century. We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less. Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes. Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens. As climate change exposes mammals to novel environmental conditions, predicting changes in animal behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat. We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

# other refs

* @gerlich_species_2025: effects of temperature on fly movement in the arctic
* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
* changes in temperature change phelogy and movement behaviour as a consequence @gerlich_species_2025
* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
\clearpage

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioural changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023]. As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022].

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where large mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.

Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021].

Changes in animal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

Our ability to respond changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in western Canada will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically, and appropriate measures of uncertainty require careful and robust modeling [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].

*Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.*

@@

As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance. In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature. Consequently, the behavioural changes we present should be interpreted carefully. While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct. Additionally, it is important to stress the distinction between habitat selection and quality. While animals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].

The diversity in predicted changes in movement behaviour by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-rss-cc) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. For example, the width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025]. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

@@

### *Canis lupus*

### *Cervus canadensis*

### *Oreamnos americanus*

### *Puma concolor*

attraction for disconnected patches of high elevation could lead to fragmentation or traps

### *Rangifer tarandus* (boreal)

Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the animals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive

### *Rangifer tarandus* (s. mountain)

HRSFs do not account for migratory behaviour due to reproduction or predation risk

### *Ursus arctos horribilis*

## Considerations for future studies

@@

While the `ctmm` movement models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency. This dependency is because coarsely-sampled movement data contains information on large-scale movements (range crossings, migrations) but not fine-scale movements, including first-order habitat selection [@johnson_comparison_1980]. Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale. @nathan_big-data_2022 provides additional examples of how the frequency of location data affects the results that can be inferred by modeling movement behaviour. Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`). Ideally, sampling schedules should be fine enough to reconstruct animals' movement movement paths. Good estimates of an animal's speed and trajectory require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction. What constitutes a change in direction depends on what movement scale one is investigating. Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second-order and third-order spatial selection. While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales, the model estimates are not independent of the scale of the sampling frequency.

When landscapes are relatively predictable across years, a larger number of sampled individuals is likely preferable over the duration of each tracking period. The greater number of tracked animals allows one to quantify the variance across individuals, including the range of conditions and environments that individual animals are in. A good estimate of inter-individual variance provides better coefficient estimates along with more appropriate measures of uncertainty. However, when conditions across years are stochastic, multi-year telemetries allow one to better estimate inter-annual variation without conflating it with inter-individual differences. Given the widespread, rapid, and accelerating changes across many habitats (particulartly at high elevations and at high latitudes), we suggest researchers focus on long-term telemetry datasets to quantify the effects of intra-annal variability while increasing the ranges of temperatures each individual is observed in. Longer observational periods can also improve the chances of observing different community dynamics, such as shifts in predator-prey dinamics.

Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour. Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities. However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common. Accurately quantifying the nonlinear effects of temperature on when, how, and where mammals move requires careful data collection and model design. We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behaviour of other species, including smaller mammals, more elusive species, and non-mammal animals. Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022]. While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables. Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study. Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity. However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.

>>>>

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles. Consequently, it is still unclear how mammals will adapt their movement behaviour in the 21^st^ century. We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023. We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century. We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less. Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes. Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens. As climate change exposes mammals to novel environmental conditions, predicting changes in mammalian behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat. We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and extreme temperatures can reduce movement rates and cause physiological damage [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioural changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

@@

Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023]. As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022;  @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

@@

\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, including how they alter seasonal and daily cycles in movement behavior. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are increasingly frequently exposed to extreme and novel conditions.

***Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour. Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities. However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common. Accurately quantifying the nonlinear effects of temperature on how and where mammals move requires careful data collection and model design. We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection. Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022]. While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables. Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study. Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity. However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.***

Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At warm temperatures, mammals overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affect mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021].

Changes in mammal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

***Protecting a third of the world's ecosystems in accordance with the Kunming-Montreal Global Biodiversity Framework [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].***

@@

As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance. In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature. Consequently, the behavioural changes we present should be interpreted carefully. While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct. Additionally, it is important to stress the distinction between habitat selection and quality. While individuals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] and as we actively prepare for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].

***...*** spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025]. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and mammals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some individuals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased risk of extirpation or extiction [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

@@

* Cougars: attraction for disconnected patches of high elevation could lead to fragmentation or traps

* boreal caribou: Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the individuals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive

* sm caribou: HRSFs do not account for migratory behaviour due to reproduction or predation risk

## Study limitation and considerations for future studies

@@

While the `ctmm` movement models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency. This dependency is because coarsely-sampled movement data contains information on large-scale movements (range crossings, migrations) but not fine-scale movements [@nathan_big-data_2022; @noonan_scale-insensitive_2019], including first-order habitat selection [@johnson_comparison_1980]. Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale. Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals. Ideally, sampling schedules should be fine enough to reconstruct individuals' movement paths. ***make data useful for future studies. recognize cost of GPS and suggest IoT [@wild_multi-species_2023], although station coverage can be a limitation [@denicola_monitoring_2025].*** Good estimates of an animal's speed and trajectory require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction. What constitutes a change in direction depends on what movement scale one is investigating. Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second-order and third-order spatial selection. While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales [@noonan_scale-insensitive_2019], the model estimates are not independent of the scale of the sampling frequency [@denicola_are_2025]. When tracking data are too coarse for accurate speed estimation, diffusion rates (measured in area per unit time) may be a useful alternative for quantifying animals' exploration rates [@denicola_are_2025]. ***also need better way to distinguish between sates. HMMs are an option, but there are limitations.***

# Other refs

* @eum_potential_2023 "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C"
* climate change threatens biodiversity of marine mammals: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
